---
title: "Cleaning the Lung function data"
output: 
  html_document: 
    highlight: tango
    theme: flatly
---

```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(knitr); library(rmarkdown);
opts_chunk$set(warning=FALSE, message=FALSE) 
```

# Import the raw data

Original data set did not have variable names included. Manually add them. 
```{r}
fev <- read.delim("Lungraw.txt", sep=" ", header=FALSE, stringsAsFactors = FALSE)
names(fev) <- c("ID", "AREA", "FSEX", "FAGE", "FHEIGHT", "FWEIGHT", "FFVC", "FFEV1", 
                            "MSEX", "MAGE", "MHEIGHT", "MWEIGHT", "MFVC", "MFEV1", 
                            "OCSEX", "OCAGE", "OCHEIGHT", "OCWEIGHT", "OCFVC", "OCFEV1",
                            "MCSEX", "MCAGE", "MCHEIGHT", "MCWEIGHT", "MCFVC", "MCFEV1",
                            "YCSEX", "YCAGE", "YCHEIGHT", "YCWEIGHT", "YCFVC", "YCFEV1")
```

Use `str(fev)` to check to see if all the variables are the correct data types. Several numeric variables
are being treated as character. Fix these. [Link to the Stack Overflow post that I googled to find a solution](http://stackoverflow.com/questions/22772279/converting-multiple-columns-from-character-to-numeric-format-in-r)
```{r}
str(fev)
cols.num <- c("MCSEX", "MCAGE", "MCHEIGHT", "MCWEIGHT", "MCFVC", "MCFEV1",
              "YCSEX", "YCAGE", "YCHEIGHT", "YCWEIGHT", "YCFVC", "YCFEV1")
fev[cols.num] <- sapply(fev[cols.num],as.numeric)
str(fev)
```


The FEV1 measurements were erroneously recoded in the data set at 100x their real value. 
```{r}
fev[c("FFEV1", "MFEV1", "OCFEV1", "MCFEV1", "YCFEV1")] <- fev[c("FFEV1", "MFEV1", "OCFEV1", "MCFEV1", "YCFEV1")]/100
head(fev$FFEV1)
```

# Save the clean data as a tab-delimited text file. 
```{r}
date <- format(Sys.Date(), "%m%d%y")
filename <- paste("Lung_", date, ".txt", sep="")
write.table(fev, filename, sep="\t", row.names=FALSE)
```

 
# Session Info
This document was compiled on `r Sys.time()` and with the following system information:
```{r}
sessionInfo()
```
