---
title: 'Raw Data'
author: ''
date: '`r Sys.time()`'
weight: 2
---

```{r include = FALSE}
library(dplyr)
library(readxl)
knitr::opts_chunk$set(echo = TRUE)
```

These data sets are used in courses that emphasize data management as well as analysis including but not limited to MATH 456 (Applied Statistics II), MATH 385 (Intro to Data Science), 485 (Advanced Data Science Topics), and MATH 615 (Graduate Statistical Methods).

All data sets listed are in their raw format. Codebooks and data management
files are supplied as available. The **Updated** column represents the
last time the data management file was edited. You should ensure that
you always have the most up to date data management file. 

```{r include = FALSE}
rawdat <- read_excel("Datasets.xlsx")

# location of files
cdat  <- rawdat %>% mutate(dataloc = filename, 
                           codeloc = codebook, 
                           dmloc   = dm,
                           Updated = format(file.info(dm)$mtime, "%m-%d-%y"))

# Add link to the data file, codebook, data management code file
cdat2 <- cdat %>% mutate(File = paste('<a href="', cdat$dataloc, '">', cdat$filename, '</a>', sep=""), 
                         Codebook = ifelse(is.na(codebook), "-", paste('<a href="', codeloc, '">', codebook, '</a>', sep="")), 
                         `Data Management` = ifelse(is.na(dm), "-", paste('<a href="', dmloc, '">', dm, '</a>', sep="")), 
                         Notes = ifelse(is.na(Notes), "-",Notes),
                         Updated = ifelse(is.na(Updated), "-",Updated))

# Select vars to display
clean <- cdat2 %>% select(Data, File, Codebook, `Data Management`, Updated, Notes)

```

<!-- unholy hack to make following two tables less wide and the same wide -->
<style type="text/css">
table {
   max-width: 50%;
}
</style>

```{r results = 'asis', echo = FALSE}
knitr::kable(clean)
```
